NumPy---Array Programming in Python

### 2020-01-11 body.tex
Minor revisions

Two different Python array packages existed before NumPy.
The Numeric package began in the mid-90s and provided an array object in
Python, written in C, and linking to standard fast implementations of linear
algebra.
Around 2000, the Space Telescope Science Institute (STScI) software group wrote
a re-implementation of much of Numeric, called NumArray, to support their work
on large memory-mapped arrays and arrays of mixed data type
records \cite{STScI-slither}.
This briefly caused the two communities to diverge, until
2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
and NumArray \cite{oliphant2006guide}.

@@

The NumPy array is a data structure that efficiently stores and accesses $n$-dimensional arrays, also known as tensors \cite{vanderwalt2011numpy}, that enables a wide variety of scientific computation.  It consists of a pointer to memory, along with meta-data used to interpret the data stored there, notably {\em data type}, {\em shape}, and {\em strides} (see Figure~\ref{fig:array-concepts}).

@@

Indexing an array returns single elements, sub-arrays, or elements that satisfy a specific condition.  Arrays can even be indexed using other arrays (see Figure~\ref{fig:array-concepts}).  Wherever possible, indexing that retrieves a subarray returns a {\em view} on the original array, such that data is shared between the two arrays.  This provides a powerful way to operate on subsets of array data, while limiting memory usage.

A powerful array feature that affects the way NumPy indexes and combines arrays, is {\em broadcasting}. When performing an operation (such as addition) on two arrays, one intuitively expect that these arrays should be required to have the same shape.  However, through broadcasting, NumPy allows the dimensions to differ, while still producing results that appeal to intuition.  A trivial example is the addition of a scalar value to an array, but it also generalizes to more complex examples such as scaling each column of an array, or generating a grid of coordinates.  In broadcasting, one or both arrays are virtually duplicated (that is, without copying any data in memory), so that the shapes of the operands match (see Figure~\ref{fig:array-concepts}).  Broadcasting is also applied whenever an array is indexed using arrays of indices.% (see Figure~\ref{fig:broadcasting}).

@@

%(e.g., summation across columns, see Figure~\ref{fig:reductions}) products, and %so forth, are
%defined.

@@

Many of these functions, such as summation, support reductions: aggregating results across over or more dimension of the array.  For example, summing a $D$-dimensional array over $p$ axes results in a $D-p$-dimensional array, as shown in Figure~\ref{fig:reductions}.

@@

Users predominantly interact with NumPy arrays using {\em indexing} (to access sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and, for matrix multiplication, $@$), as well as {\em utility functions}; together, these provide an easily readable, expressive, high-level API for array programming.  Exposing these array programming primitives in Python---an interpreted language---creates an ideal environment for exploratory data analysis \cite{perez2007ipython} where users may inspect, manipulate, and visualize their data.

\fixme{This end of paragraph needs fixing up}Instead of focusing on programming, users can bring their attention to understanding and interacting with their data.

% + interactivity
% users tend to "live in their data",
% interactivity and powerful, expressive syntax ... inspect, feel their way forward, experiment
% collect these into imperative or functional programs ....
% not that they never plan, often do ... but the ability to stopping "programming"
% and experiment is a key factor in why people use numpy ....

\section*{Changing computational landscape}

As Python's popularity for scientific computing, data science, and machine
learning increased, many new challenges and use cases arose.

\fixme{The idea here is something like:  classic numpy (as described above)
was built to solve previous problems; new problems have arisen as
the computational landscape has changed; new hardware, new languages,
new problems; as a result np's role has changed; it is a foundational
tool widely-used not a specialist tool used by scientist developers;
many array-like libraries for various uses and np is serving as
a standard)}

@@

- distributed computing

@@

- massive explosion in data science, machine learning, and artificial intelligence

NumPy also went from a small community of mostly expert users to
one of the most widely used foundations of the scientific Python
ecosystem.
Now the user base is very different....
% maybe worth a sentence or two about how we are working on web and
% user docs to make it easier for new users to get started...

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.8\textwidth]{static/sketches/duck-arrays}
  \caption{\textbf{Duck Arrays.} \fixme{Examples of so-called ``duck arrays'': arrays
  implemented by another library that behave like NumPy arrays, and
  support NumPy operations.  Through protocols, it is possible to have
  a call such as $\cos(\cdot)$ succeed on each of these.}}
  \label{fig:duck-arrays}
\end{figure*}

NumPy's API has become ubiquitous in scientific computation, but NumPy itself can never hope to satisfy every single specialized need of the community.  The advent of new technologies, such as GPUs, create opportunities that should be exploited, but are out of scope for NumPy development.  Still, the widespread adoption of the NumPy API is valuable, as it lowers the barrier to entry and provides stability to the community.  What can NumPy do to support the use of its API without blocking innovation?  We believe that improving interoperability by having NumPy be a central coordinator among array objects is a good way to balance stability while supporting growth of a larger surrounding ecosystem of specialized solutions and new tools.

Primarily, there exist two types of Python array objects: subclasses of NumPy arrays, and arrays that mimic NumPy arrays but are fundamentally different.

Subclasses of NumPy often exist due to difficulty in constructing richer data types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint}, geometrical objects \cite{pygeos}, and missing numbers.  To improve {\em extensibility}, we are currently overhauling the data type system to make it easier to implement data types both in Python and C, and to support these and other applications.

@@

To facilitate {\em interoperability}, the NumPy team is defining so-called ``protocols'' (or contracts of operation), that allow for these arrays to be passed to NumPy functions.  NumPy, in turn, dispatches operations to the originating library, as required.

\url{https://numpy.org/neps/nep-0037-array-module.html}

@@

The creation of NumPy spurred renewed development in the larger scientific
Python ecosystem and heralded the current era of wide-spread use of Python for
scientific computing.
There were several factors that allowed this rapid growth
and successful development.

Hugunin identified the first factor in his initial description of the Numeric
package \cite{Hugunin-whitepaper}.  Numerical computing is usually one
component of a larger programming task, therefore:
\begin{quote}
    Rather than trying to retrofit an existing numerical language to support
    the wealth of features found in a powerful, modern, general-purpose
    programming language, it makes much more sense to attack the problem from
    the other direction and add the features of a powerful numerical
    programming language to Python.
\end{quote}
Python is already a good choice for many standard programming tasks such as
cleaning data, interacting with web resources and parsing text.  It has a wide
range of libraries for many different tasks. Adding fast array operations and
linear algebra allows the scientist to do all their work with within the same
language---and one that has the advantage of being famously easy to learn and
teach.

The second factor flows from Python's nature as an open-source language,
embedded within the open-source community.  Python attracts developers who like
to build and contribute.  As a result, NumPy has always been a library that was
developed by scientists in order to do their work.  Commercial specialist
numerical languages can encourage consumer-users, who pick up a tool and use
it, but do not expect to contribute to development. %<-sentence a bit tough to parse.
In contrast, NumPy often
attracts user-developers, who want to work with NumPy, and are prepared to fix
and extend the library as they work. User-developers have been the fundamental
force in developing NumPy, and shaping it into a library that has been refined
in the furnace of practical scientific work.

The third factor is much like the second: 
%a culture in Python that concentrates
%on high quality of code and distribution.  Because Python is open-source, it has
%a strong culture of contribution, and therefore, of using tools and process to
%improve collaborative software development, such as distributed version
%control, code review, and automated testing.  These have allowed the project to
%scale safely as it attracted more users and developers, even though these
%developers rarely came to the project with training in software engineering.
the project has strong culture of using software-engineering practice to
improve collaboration and reduce error \cite{millman2014developing}.
The NumPy team was early in adopting distributed
revision control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for every proposed
change to NumPy.
The project has comprehensive, high-quality documentation,
integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

Lastly, we believe NumPy has benefited greatly from a well-chosen application
programming interface (API).  Scientific and numerical programming needs to be
fast, and scale to very large datasets.  The NumPy API defines a simple wrapper
for data in memory so it can be represented as a one- or multi-dimensional
array.  The simplicity of this wrapper has made it very successful as a
standard way of representing arrays in memory, and has made it relatively easy
for other libraries to develop fast and memory-efficient compiled code, usually
in C or Fortran, that can manipulate these arrays and pass them back to Python.
This has been a large factor in the growth of numerical computing libraries
around NumPy and Python.

>>>>

Two Python array packages existed before NumPy.
The Numeric package began in the mid-90s and provided an array object in
Python, written in C, and linking to standard fast implementations of linear
algebra.
Around 2000, the Space Telescope Science Institute (STScI) software group wrote
a re-implementation of much of Numeric, called NumArray, to support their work
on large memory-mapped arrays and arrays of mixed data type
records \cite{STScI-slither}.
This briefly caused the two communities to diverge, until
2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
and NumArray \cite{oliphant2006guide}.

@@

The NumPy array is a data structure that efficiently stores and accesses
$n$-dimensional arrays \cite{vanderwalt2011numpy}, also known as tensors, that
enables a wide variety of scientific computation.
It consists of a pointer to memory, along with meta-data used to interpret the
data stored there, notably {\em data type}, {\em shape}, and {\em strides}
(Fig.~\ref{fig:array-concepts}a).

@@

Indexing an array returns single elements, sub-arrays, or elements that satisfy
a specific condition (Fig.~\ref{fig:array-concepts}b).
Arrays can even be indexed using other arrays (Fig.~\ref{fig:array-concepts}d).
Wherever possible, indexing that retrieves a subarray returns a {\em view} on
the original array, such that data is shared between the two arrays.
This provides a powerful way to operate on subsets of array data, while
limiting memory usage.

A powerful array feature that affects the way NumPy indexes and combines
arrays, is {\em broadcasting}.
When performing an operation (such as addition) on two arrays, one intuitively
expect that these arrays should be required to have the same shape.
However, through broadcasting, NumPy allows the dimensions to differ, while
still producing results that appeal to intuition.
A trivial example is the addition of a scalar value to an array, but it also
generalizes to more complex examples such as scaling each column of an array,
or generating a grid of coordinates.
In broadcasting, one or both arrays are virtually duplicated (that is, without
copying any data in memory), so that the shapes of the operands match
(Fig.~\ref{fig:array-concepts}c).
Broadcasting is also applied whenever an array is indexed using arrays of
indices (Fig.~\ref{fig:array-concepts}d).

@@

%(e.g., summation across columns, see Fig.~\ref{fig:reductions}) products, and %so forth, are
%defined.

@@

Many of these functions, such as summation, support reductions: aggregating
results across over or more dimension of the array.
For example, summing a $D$-dimensional array over $p$ axes results in a
$D-p$-dimensional array (Fig.~\ref{fig:reductions}).

@@

Users predominantly interact with NumPy arrays using {\em indexing} (to access
sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$,
and, for matrix multiplication, $@$), as well as {\em utility functions};
together, these provide an easily readable, expressive, high-level API for
array programming.
Exposing these array programming primitives in Python---an interpreted
language---creates an ideal environment for exploratory data analysis where
users may inspect, manipulate, and visualize their data
\cite{perez2007ipython}.

\section*{The proliferation of arrays}

\begin{figure*}
  \centering
  \includegraphics[width=0.8\textwidth]{static/sketches/duck-arrays}
  \caption{\textbf{Proliferation of Arrays.} Many ...}
  \label{fig:array-proliferation}
\end{figure*}

(Fig.~\ref{fig:array-proliferation})

- distributed arrays

@@

- computation graph

- sparse arrays

@@

- massive explosion in data science, machine learning, and artificial intelligence (tensors)

@@

NumPy's API has become ubiquitous in scientific computation, but NumPy itself
can never hope to satisfy every single specialized need of the community.
The advent of new technologies, such as GPUs, create opportunities that should be
exploited, but are out of scope for NumPy development.
Still, the widespread adoption of the NumPy API is valuable, as it lowers the
barrier to entry and provides stability to the community.
What can NumPy do to support the use of its API without blocking innovation?
We believe that improving interoperability by having NumPy be a central
coordinator among array objects is a good way to balance stability while
supporting growth of a larger surrounding ecosystem of specialized solutions
and new tools.
Primarily, there exist two types of Python array objects: subclasses of NumPy
arrays, and arrays that mimic NumPy arrays but are fundamentally different.

Subclasses of NumPy often exist due to difficulty in constructing richer data
types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint},
geometrical objects \cite{pygeos}, and missing numbers.
To improve \emph{extensibility}, we are currently overhauling the data type
system to make it easier to implement data types both in Python and C, and to
support these and other applications.

@@

To facilitate \emph{interoperability}, the NumPy team is defining so-called
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions.
NumPy, in turn, dispatches operations to the originating library, as required.

%\url{https://numpy.org/neps/nep-0037-array-module.html}

@@

NumPy spurred the development of the larger scientific Python ecosystem and
heralded the current era of wide-spread use of Python for scientific computing.
There were several factors that allowed this rapid growth and successful
development.

Python is already a good choice for many standard programming tasks such as
cleaning data, interacting with web resources and parsing text.
It has a wide range of libraries for many different tasks.
Adding fast array operations and linear algebra allows the scientist to do all
their work with within the same language---and one that has the advantage of
being famously easy to learn and teach.

NumPy has strong culture of using software-engineering practice to
improve collaboration and reduce error \cite{millman2014developing}.
The NumPy team was early in adopting distributed revision control and code
review to improve collaboration on code, and continuous testing that runs a
large battery of automated tests for every proposed change to NumPy.
The project has comprehensive, high-quality documentation, integrated with the
source code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

Lastly, NumPy has benefited greatly from a well-chosen API.
Scientific and numerical programming needs to be fast, and scale to very large
datasets.
The NumPy API defines a simple wrapper for data in memory so it can be
represented as a one- or multi-dimensional array.
The simplicity of this wrapper has made it very successful as a standard way of
representing arrays in memory, and has made it relatively easy for other
libraries to develop fast and memory-efficient compiled code, usually in C or
Fortran, that can manipulate these arrays and pass them back to Python.
This has been a large factor in the growth of numerical computing libraries
around NumPy and Python.

### 2020-01-11 summary.tex
Minor revisions

% - 2-3 sentences of basic-level introduction to the field;
Array programming provides powerful notation for accessing,
manipulating, and operating on data in vectors, matrices, and
higher-dimensional arrays \cite{iverson1980notation}.
% - a brief account of the background and rationale of the work;
NumPy is the fundamental array programming library for the Python ecosystem
\cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
It has been actively developed for over two decades and has more than 850
contributors, nearly 300,000 dependent repositories on GitHub, and millions of
downloads per year.
It plays an essential role in research analysis pipelines in fields as
diverse as physics, biology, astronomy, neuroscience, material science,
engineering, and chemistry.
For example, in astronomy, NumPy was fundamental to discovering gravitational
waves\cite{pycbc}, the first imaging of a black hole\cite{eht-imaging}, and
continues underpinning the future exploration of our
universe\cite{jenness2018lsst}.
It is one of the most commonly used machine learning tools among
enterprises \cite{451report2018}.
% - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
Here we show how a small group of students and scientists created this
popular tool, explain the core ideas that helped make it it successful,
and highlight recent development to better serve new use cases
and future improvements.
% - 2-3 sentences putting the main findings into general context so it is
%   clear how the results described in the paper have moved the field forwards.
NumPy continues to evolve to meet the needs of its community, to remain
relevant in an ever-changing computing landscape, and to be a fundamental
pillar of support for the next decade of scientific computing, data science,
and machine learning in Python.

>>>>

% - 2-3 sentences of basic-level introduction to the field;
Array programming provides powerful notation for accessing,
manipulating, and operating on data in vectors, matrices, and
higher-dimensional arrays \cite{iverson1980notation}.
% - a brief account of the background and rationale of the work;
NumPy is the fundamental array programming library for the Python programming language
\cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
It has been actively developed for over two decades and has more than 850
contributors, nearly 300,000 dependent repositories on GitHub, and millions of
downloads per year.
It plays an essential role in research analysis pipelines in fields as
diverse as physics, biology, astronomy, neuroscience, material science,
engineering, and chemistry.
For example, in astronomy, NumPy was fundamental to discovering gravitational
waves\cite{pycbc}, the first imaging of a black hole\cite{eht-imaging}, and
continues underpinning the future exploration of our
universe\cite{jenness2018lsst}.
It is one of the most commonly used machine learning tools among
enterprises \cite{451report2018}.
% - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
Here we show how a few core ideas lead to a simple and powerful array syntax.
% - 2-3 sentences putting the main findings into general context so it is
%   clear how the results described in the paper have moved the field forwards.
NumPy arrays are so popular that several project have developed their own
NumPy-like interface and array objects targeting audiences with specialized
needs.
Because of this, NumPy plays a unique role in enabling disparate technologies
to interact and exchange.
%NumPy continues to evolve to meet the needs of its community, to remain
%relevant in an ever-changing computing landscape, and to be a fundamental
%pillar of support for the next decade of scientific computing, data science,
%and machine learning in Python.
